SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2013-01-01 paper/paper.tex
paper: updates.

class Arbitrary a => SubTypes a where
  size    :: a -> Size
  index   :: a -> Idx -> Maybe SubVal
  replace :: a -> Idx -> SubVal -> a
  constr  :: a -> String
  constrs :: a -> [String]
\end{code}
%
\noindent
The \ttp{SubTypes} type class requires QuickCheck's \ttp{Arbitrary} as a
super-class.  \ttp{SubTypes} has five methods.  The \ttp{size} method returns
the size of a value, \ttp{index} returns a sub-value at an breadth-first index
in a value, \ttp{replace} replaces a sub-value at a particular focus,
\ttp{constr} returns the constructor tagging the value, and \ttp{constrs}
returns a list of unique identifiers for the constructors associated with each
variant of the type.

For example, consider a binary tree type:
%
\begin{code}
data Tree = L | B Tree Tree
\end{code}
%
\noindent
and the value \ttp{tree}, labeled with the indexes of its sub-values
%
\begin{code}
tree = B\sub{0} (B\sub{1} L\sub{3}
             (B\sub{4} L\sub{6} L\sub{8}))
          (B\sub{2} L\sub{5} L\sub{7})
\end{code}
%
\noindent
Here are example applications of \ttp{SubTypes} methods; in the following, we
show the indexes with respect to the \ttp{tree} value:
%
\begin{code}
size tree = 9

@@

constr  tree = ["B"]
constrs tree = ["B", "L"]
constrs L    = ["B", "L"]
\end{code}
%
\noindent
We will sometimes refer to the unwrapped value returned by \ttp{index a i} as
the \emph{$i$th sub-value of $a$}, so for example, \ttp{B\sub{4} L\sub{6}
  L\sub{8}} is the 4th sub-value of \ttp{tree}.  An invariant of \ttp{index} is
that for any value \ttp{a}, consider the least $i \geq 0$ such that $$\ttp{index
  a i} = \ttp{Nothing}$$ For all $0 \leq j < i$, $$\ttp{index a j} \neq
\ttp{Nothing}$$ We use this invariant as a termination case in recursive
algorithms over the sub-values of a value.

@@

Finally, while \ttp{constr} returns a string representation of the constructor
associated with the specific variant passed to it, \ttp{constrs} ignores the
value and returns the same list of constructors for any value from the type.
(In our implementation, \ttp{constrs} depends on GHC Generics~\cite{generics},
which we describe in Section~\ref{sec:implementation}.  For simplicity, we omit
here Generics-specific super-class constraints on the \ttp{SubTypes} class.)

@@

\subsubsection{Base types}

@@

\todo{Add discussion here, note may be overapprox, since testing}

\subsection{Automated Precondition Strengthening}

The extrapolation algorithm generalizes a counterexample, but in the
``neighborhood'' of he original counterexample.  In particular, all
generalizations are from the same variant as the original counterexample.  To
help the programmer in the generalization step, we would also like a way to test
the property again, ensuring we get counterexamples (if they exist) outside of
the neighborhood of the original one.

@@

Figure~\ref{fig:cex-gen} illustrates the situation in which we have some values
and a property in the case that some values fail the property of the form
(\ttp{pre ==> prop}).  Our goal is to find additional counterexamples in the
shaded region, in which \ttp{pre} is satisfied, \ttp{prop} is not satisfied, and
the counterexample is not already characterized by a universal sub-value
generalization formula.

@@

@@

\todo{xmonad}
\todo{use SC to test itself?}

@@

>>>>

class Arbitrary a => SubTypes a where
  size     :: a -> Size
  index    :: a -> Idx -> Maybe SubVal
  replace  :: a -> Idx -> SubVal -> a
  constr   :: a -> String
  constrs  :: a -> [String]
  baseType :: a -> Bool
\end{code}
%
\noindent
The \ttp{SubTypes} type class requires QuickCheck's \ttp{Arbitrary} as a
super-class.  \ttp{SubTypes} has the following methods:
\begin{itemize}
\item \ttp{size} returns the size of a value,
\item \ttp{index} returns a sub-value at an breadth-first index in a value,
\item \ttp{replace} replaces a sub-value at a particular focus,
\item \ttp{constr} returns the constructor tagging the value,
\item \ttp{constrs} returns a list of unique identifiers for the constructors
  associated with each variant of the type,
\item \ttp{baseType} returns whether the type of the value is an ``interesting
  type'', which generally means it has structure (e.g., a sum or product type).
  We discuss the purpose of base types in Section~\ref{sec:base}.
\end{itemize}

\noindent
For example, consider a binary tree type:
%
\begin{code}
data Tree = L | B Tree Tree
\end{code}
%
\noindent
and the value \ttp{tree}, labeled with the indexes of its sub-values
%
\begin{code}
tree = B\sub{0} (B\sub{1} L\sub{3}
             (B\sub{4} L\sub{6} L\sub{8}))
          (B\sub{2} L\sub{5} L\sub{7})
\end{code}
%
\noindent
Here are example applications of \ttp{SubTypes} methods; in the following, we
show the indexes with respect to the \ttp{tree} value:
%
\begin{code}
size tree = 9

@@

constr  tree = ["B"]
constrs tree = ["B", "L"]
constrs L    = ["B", "L"]

baseType (3 :: Int) = True
baseType L          = False
\end{code}
%
\noindent
We will sometimes refer to the unwrapped value returned by \ttp{index a i} as
the \emph{$i$th sub-value of $a$}, so for example, \ttp{B\sub{4} L\sub{6}
  L\sub{8}} is the 4th sub-value of \ttp{tree}.  An invariant of \ttp{index} is
that for any value \ttp{a}, consider the least $i \geq 0$ such that $$\ttp{index
  a i} = \ttp{Nothing}$$ For all $0 \leq j < i$, $$\ttp{index a j} \neq
\ttp{Nothing}$$ We use this invariant as a termination case in recursive
algorithms over the sub-values of a value.

@@

While \ttp{constr} returns a string representation of the constructor
associated with the specific variant passed to it, \ttp{constrs} ignores the
value and returns the same list of constructors for any value from the type.
(In our implementation, \ttp{constrs} depends on GHC Generics~\cite{generics},
which we describe in Section~\ref{sec:implementation}.  For simplicity, we omit
here Generics-specific super-class constraints on the \ttp{SubTypes} class.)

Finally, \ttp{baseType} is true for our tree type but false for a structureless
type, such as \ttp{Int}.

@@

\subsubsection{Base types}\label{sec:base}

@@

The extrapolation algorithm is unsound insofar as it extrapolations from a set
of counterexamples to a universal claim, similar to QuickSpec~\cite{qs}.  By
tuning the parameters, the risk of an unsound generalization is reduced.

\subsection{Automated Precondition Strengthening}
The extrapolation algorithm generalizes a counterexample, but in the
``neighborhood'' of he original counterexample.  In particular, all
generalizations are from the same variant as the original counterexample.  To
help the programmer in the generalization step, we would also like a way to test
the property again, ensuring we get counterexamples (if they exist) outside of
the neighborhood of the original one.

@@

Figure~\ref{fig:cex-gen} illustrates the situation in which we have some values
and a property in the case that some values fail the property of the form
(\ttp{pre ==> prop}).  Our goal is to find additional counterexamples in the
shaded region, in which \ttp{pre} is satisfied, \ttp{prop} is not satisfied, and
the counterexample is not already characterized by a universal sub-value
generalization formula.  Iterating, we increasingly cover the shaded region
with new formulas until no more uncovered counterexamples exist or the testing
tool fails to find new counterexamples.

\begin{figure}

  \caption{Shape matching algorithm.}
  \label{fig:matches}
\end{figure}

@@

Unlike universal sub-value generalization, existential sub-value generalization
is sound.  The existential claim is only that for each variant, there exists at
least one counterexample, so inductive reasoning is not required.

@@

\todo{xmonad}
\todo{use SC to test itself?}
\todo{experiments? scale vs. broad data, linear data?}

@@

\todo{Use Bayes' rule to do extrapolation?}
